PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-03-03 reviewer_response.tex

\subsection*{Reviewer 3}

\reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure. Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose.\textbf{ The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.}\\
      For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''. There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG. There are two primary differences in these existing tools: those which generate a graph at the nucleotide level with nodes representing basepairs (sometimes nonbranching runs of nodes are collapsed) and those where nodes represent larger features such as genes.\\
      PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.}

\response{
      We apologize if our previous discussion of the literature was incomplete, it was not our intention to downplay prior work.
      We thank the reviewer for highlighting additional important work and aspects that we previously didn't discuss.
      We have rewritten the part in the introduction to reflect these the work by these and other authors and added a few sentences to the discussion to further highlight previous and alternative approaches. We hope that these edits address the concerns of the reviewer.
      % SURVEY OF THE LITERATURE
      % PGG have been used to:
      % - improve annotations [sutton 2021] + [ggcaller]
      % MERTIS OF OUR METHOD
      % - de Bruijn graphs represent SNPs and large scale variation in the same way, as edges in the graph. We instead have this scale separation and encode nucleotide-level variation in alignments, and structural changes in the graph
      % - PGG based on genes are sensitive to annotation errors. We are based on homology.
      % - Scalable algorithm.
}

\Marco{
      Finish literature review.
      Add part on graph-based methods.
      In the discussion add edge cases.
      Cite all the ones suggested by the reviewer.
}

\reviewer{R3}{An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists. The mechanism for this is obvious but providing a useful user interface might be tedious.}
% \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
% \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
\response{We thank the reviewer for this important suggestion. We appreciate the need to include annotation in our pangenome graph structure. We created a script to decorate each annotation in a \texttt{gff} file with information on the pancontig location. The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}. We plan on extending this in the future to use directly the pangenome graph as a reference and add annotations as features of pancontigs.
      % TODO: mention this in the text.
      % TODO: make the repository public.
}

\reviewer{R3}{This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging. Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017).'' appears to be incomplete at best. De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.
      % \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained. Every SNP can generate a branching event. While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes. I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\
      One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
      \begin{itemize}
            \item \cite{angiuoli2011improving}
            \item \cite{angiuoli2011mugsy}
      \end{itemize}
      While the underlying approach is quite different, a very similar PGG to PanGraph is produced by a different method with similar nodes represented by multiple sequence alignments and edges with the same meaning:
      \begin{itemize}
            \item \cite{sutton2021pan}
            \item \cite{chan2015novel}
      \end{itemize}}
\responsedraft{}
\Marco{Copy form above and extend the answer}

@@

However, as the reviewer points out, in the case of distantly related paralogs, such a grouping might be undesirable as it introduces unneeded complications in the graph. Currently \textit{PanGraph} supports two different ways of partially controlling paralog splitting.\\
The first one is based on sequence divergence. By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
The second method relies instead on context. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. These duplicated pancontigs are then split and absorbed in their context.\\

In the future we plan on providing a dedicated command to perform a more robust form of paralag splitting, which combines both sequence divergence information and context information. This command would build a phylogenetic tree from the alignment of each duplicated pancontig, to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context. This functionality is not however not yet implemented in the current version of \textit{PanGraph}.\\

We now make this explicit in SI sect.~{\SIalgo}D, that now contains these considerations.

% TODO: if this is sensible add a paragraph in SI about paralog splitting
% TODO: confirm with Nick the paralog split algorithm.
}
\Marco{Summarize, make more to the point. Say that this is now better discussed in SI.}

@@

\reviewer{R3}{How are pancontig consensus sequences determined and updated? I could not find any details about this. This is not a trivial operation. My reading of the article seems to imply that a complete multiple sequence alignment is not done for each merger. Does the consensus sequence degenerate iteratively with mergers? For the simulated data how accurate are the final pancontig consensus sequences?}
% \Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address. Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
% \Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI. The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks. Then consensus of the full alignment is re-evaluated. This does not solve the multiple alignment problem during constructions and there can be degeneration on where indels are mapped, but the final graph can be eventually polished. I don't think we ever quantified how much the consensus sequence degenerates during the iterations. Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
\response{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer. To this end we added section {\SIalgo} in the SI (see ``graph merging''), with a detailed explanation of how the merging procedure works.\\
      In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs. We avoid performing a multiple sequence alignment during merging to maintain the algorithm scalable. However we do update the consensus of the pangontig at each merge using the majority rule on the merged alignment.\\
      The idea behind this approach is that the consensus sequence is needed to find long-range homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an eccessive amount of resources. However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command. This command performs a multiple sequence alignment on all the sequences in each pancontig and updates the consensus sequence accordingly. This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
      We did not quantify explicitly the divergence of consensus sequences during graph building, but we observed that, even without ``polishing'', alignments of up to hundreds of sequences look consistent.
      % TODO: quantify this in simulations if there is time?
}

@@

To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
% TODO: core genome for simulated data if there is time?
}
\Marco{Make more to the point. Data are from PanX (archived run May 2018, not the version used in the paper). Add the date. First point at the reason: core genome and not soft core. Similar problems in other pipelines. Also other, but not dramatic. Give numbers}

%     "all core": 778,
%     "all softcore": 2081,
%     "all genes": 23128,
%     "all nondupl": 18798,
%     "drop core": 822,
%     "drop softcore": 2069,
%     "drop genes": 23054,
%     "drop nondupl": 18749

\reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}
% \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
\response{
      We appreciate the importance of the point raised by the reviewer.
      In it's current version \textit{PanGraph} does not support a post-processing refinement step in which alignments of adjacent edges of pancontigs are harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones.
      We agree with the reviewer that this post-processing smoothing of the graph would be an important improvement, and we consider adding it in subsequent versions of \textit{PanGraph}.
}

\reviewer{R3}{Overall the paper is well written. I like the basic methodology and the attention to using robust efficient existing algorithms and combining them to keep efficiency in mind. I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.}
% \Marco{if the fracturing derives from a real structural property of pangenomes (i.e. ISs that really move around and split long blocks) this might be an intrinsic limit that derives from the pangenome, I wouldn't consider it a problem with PanGraph.}\\
\responsedraft{
      We thank the reviewer for their appreciation of the paper and of our methodology.\\
      We agree with their concern that the level of fragmentation of the graph constitutes a limitation in the number and diversity of genomes that can meaningfully be added in the pangenome graph.\\
      An increase in the number or diversity of isolates included in the graph will cause an increase in structural breakpoints in the graph. If subsequent structural changes occurr at a scale smaller than $\Lthr$, these will not correctly be captured by the graph topology. An excessive number of pancontigs would also increase the amount of computational resources used in the alignment.\\
      % Other than being introduced by the diversity of the dataset, fragmentation could also be introduced artificially in the merging procedure. We tested for excessive fragmentation in simulated data using partition entropy, comparing the real underlying structural partition to the one reconstructed by pangraph (not shown in the paper). We found that fragmentation was present only when the diversity of the sample was too high for the sensitivity of the aligner.\\
      Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of block size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel). Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in blocks having size close to $\Lthr$.\\
      (As soon as they are done I could put the time and block size distribution for graphs with 500/1000 E.coli.)
      % TODO: show distributions for pangenome graphs with 500/1k isolates? Mention times?
      % TODO: add this discussion in the text?
      % TODO: mention the partition entropy? This could be an argument against artificial fragmentation.
      % TODO: maybe add in the SI the block frequency distribution for different species/kernels?
}
\Marco{It is an important concern. We have investigated it in the paper, good scaling... Can be detected if wrong...}

\reviewer{R3}{An issue I have struggled with in my own work is how to represent an orthologous gene cluster (node/vertex) within the PGG when the gene sequence in some genomes has an insertion event such as an IS element within the gene. I assume for PanGraph that this would necessitate splitting the node into multiple pancontigs. How is this sustainable for species with large numbers of active IS elements? This could also be investigated for PanGrah using suitable simulated data.}
% \Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots). For the future one might want to have pangenome methods where you can explicitly pass a list of predefined sequences that should not be split. But I think this is a general problem and not something to investigate with simulations or fix here.}\\
% \Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}
\responsedraft{
      We agree with the reviewer's observation that events such as the presence of an IS in an orthologous gene cluster would split the cluster in multiple pancontigs. For datasets in which isolates have a high amount of IS (larger than $\Lthr$), this would indeed result in multiple splits in the graph structure.\\
      This is in part an explicit design choice in \textit{PanGraph}. We plan to use this tool for the study of the evolution of bacterial pangenome, and we wanted this feature of the pangenome - namely the fragmentation due to presence of the same IS in multiple contexts - to be explcitly encoded in the graph. By condensing all occurrences of such an IS in a single pancontig we also condense the information on all of the different contexts in which the said IS appears, at the cost of fragmenting the graph. We are already using this property in a work in preparation \cite{shaw_preparation}, in we compare the similarity between different contexts to obtain insights in the evolution of such sequences. This is especially important given the fact that IS often carry limited phylogenetic information, due to their limited size.
      % TODO: maybe a bit off the mark? Not necessary to talk about the work in preparation?
}
\Marco{Future work, requires further research. Will consider adding specific procedures for IS (e.g. give a list of IS sequences that could be excluded...)}

### 2023-03-04 reviewer_response.tex

\reviewer{R3}{The statement: "Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs." needs clarification. Why would these pancontigs have been split in the first place? Is there any necessary special issues with producing the merged multiple sequence alignment?}
% \Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
% \Marco{I can think of one edge case which is due to periodic boundary conditions. Initially each circular genomes is saved as a single long block. If during alignment this unique blocks is split somewhere in the middle, the resulting edge blocks might be always colinear, since they are actually adjacent parts of the sequence that were artificially split to linearize the circular genome. One wants to collaps these two parts in a single block. This should happen often at the beginning of the merging of circular genomes. But might be unnecessary detail when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
\response{Transitive pancontigs are an artifact of periodic boundary conditions when merging circular genomes, such as bacterial chromosomes. At the beginning of the procedure the circular genome is saved as a single linear pancontig. This introduces an artificial cut. When during merging this unique pancontig is further split, the flanking pancontigs could be crossing the artificial cut in a co-linear way on all paths. This step of the procedure resolves the artificial cut buy merging these pancontigs together.\\
      We think that the mention of this step of the algorithm in this part of the paper might be a bit confusing and needs more clarification. We moved and expanded the statement from section IB of the main text to SI section I (point 5 of the merging algorithm explanation).}

\reviewer{R3}{It is not clear to me what Table 1 shows. Is this published data from PanX or data from PanGraph? Either way the value of 0.7 Mbp for the E. coli core genome length seems implausibly low for high quality complete genomes. The authors should justify this. The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.}
% \Liam{0.7Mbp does seem lower than I would expect - I guess that they expect $>$2 Mbp - but it shows that obsessing about core is a bit unwise. With slightly relaxed definitions (95\% of genomes) does the core get sensibly bigger? For comparison e.g. ggCaller gets ~2,500 core genes in 100\% of n=162 \textit{E. coli} (Table 1 of \href{https://www.biorxiv.org/content/10.1101/2023.01.24.524926v1.full.pdf}{here}).}\\
% \Marco{I checked and soft core (gene in 95\% of strains) is 2Mbp in our dataset. Some of our problems might come from weird strains. E.g. the second strain with the least amount of soft-core genes is \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518}, an artificial genome that was removed from RefSeq. The strain that misses the most of the 1000 most frequent genes \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663} was also removed from RefSeq. The rest might be due to misannotations? At least one of the most common 1000 genes are missed by 87/307 strains. Interestingly in \cite{sutton2021pan} they report $\sim$700 core orthologous gene clusters for an E.coli dataset of $\sim$1000 genomes, if they do not harmonize annotations. Those increase to $\sim$1500 after refining their annotations.
% The easiest thing to do might be to just add the size of soft-core for all species? Maybe these values might also be added as dashed lines in fig. 4G for comparison with pangraph's estimation?}\\
\response{
We thank the reviewer for pointing out this issue with the E.~coli dataset.\\

Concerning the origin the data, our datasets are taken from the PanX pipeline (\url{https://pangenome.org/}). They were not included in the original publication but are part of a later update, dated March 2018. Values displayed in table I are extracted from the PanX analysis. This is now made more explicit in the text.\\

Concerning the core genome size, the unusually small value of 0.7 Mbp is due to the fact that we used a strict definition of core genes as single-copy genes that are present in every isolate. This definition is very sensitive to annotation errors.\footnote{
      For example in \cite{sutton2021pan} using a similar strict definition the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes (see table 2 in their publication), compatible with our estimate of 0.7 Mbp. They impute this to misannotations, since after refining the annotations using structural infromation from the pangenome this number increases to $\sim$1500 genes.
} If we measure the size of the soft-core genome with a frequency threshold of 95\% we find a more reasonable size of 2 Mbp. We now report both of these values in table~I.\\
A careful investigation of the dataset also revealed that a pair of records have been subsequently removed from RefSeq, namely \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518} (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}. However these are not at the origin of the small size of the core genome. In fact by removing these isolates the number of core genes only shows a very limited increase from 778 to 822, and the number of soft-core genes decreases from 2081 to 2069.\\

To make this more clear we updated part of the text in section IIB, the caption of table~I and SI sect.~{\SIdataBenchmark}A.
}
% The isolates analyzed in our papers were downloaded from RefSeq. The accession numbers are now explicitly included in the sumbission (file \texttt{accession\_numbers.csv}). These data were previously analyzed in our group with the PanX pipeline \cite{ding2018panx}. We rely on this pipeline for the clustering of homologous genes, and the values shown in table~1 results from this analysis. This is now made explicit in the caption. This table is only meant to provide some general characterization of the size and diversity of the various datasets, and PanX analysis results are not used in PanGraph.\\

% As noted by the reviewer, the 0.7 Mbp size for E.~coli core genome found by PanX is smaller than what usually reported in the literature. Estimates of E.~coli pangenome size are usually around 2000-2500 genes \cite{park2019large,horsfield2023accurate,sutton2021pan}. However these estimates often either use a soft-core definition for the core genome \cite{park2019large} or use some form of annotation refinement to mitigate misannotations \cite{horsfield2023accurate,sutton2021pan}.\\
% In our case since we are using a strict definition for core genome the discrepancy could have different origins: problematic assemblies in our datasets, misannotations or problems with gene clustering.\\
% As a first step we checked the size of the soft-core genome, defined as the set of gene clusters that are shared by at least 95\% of the strains. For our E.~coli dataset this has size of 2 Mbp, which is closer to our expectation.\\
% We then ranked strains by the amount of soft-core genes that they include, and found some problematic assemblies. Some of the isolates with the lowest fraction of soft-core genes were subsequently removed from RefSeq, for example \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518} (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}.\\
% These problematic strains however do not explain the small core-genome size that we find. For instance, over the 307 strains in our dataset, 87 are missing at least one of the most common 1000 genes. This discrepancy is however compatible with misannotations. For instance by using a strict core definition in \cite{sutton2021pan} the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes (see table 2 in their publication). This would be compatible with our estimate of 0.7 Mbp. After refining the annotations using structural infromation from a pangenome graph, the number of strict core gene clusters increases to $\sim$1500. In our PanX analysis we do not perform this harmonization and our esitmate remains biased by annotation errors.\\

% To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
% TODO: core genome for simulated data if there is time?

### 2023-03-04 main.tex
updated panx data answer

\paragraph*{Dataset characterization.}
We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus marinus} (PM). These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf. Table~\ref{table:panx-dataset} and SI Section~{\SIdataBenchmark}A).

>>>>

\paragraph*{Dataset characterization.}
We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus marinus} (PM). These data had been previously analyzed using PanX \cite{ding2018panx}. Using the PanX analysis we estimated the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf. Table~\ref{table:panx-dataset} and SI Section~{\SIdataBenchmark}A).

% We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia coli} (EC), \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Mycobacterium tuberculosis} (MT) and \textit{Prochlorococcus marinus} (PM). These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf. Table~\ref{table:panx-dataset} and SI Section~{\SIdataBenchmark}A).

### 2023-03-04 supplementary.tex
updated panx data answer

We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species: \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Prochlorococcus marinus} (PM), \textit{Mycobacterium tuberculosis} (MT) and \textit{Escherichia coli} (EC). In Table~I in the main text we report some summary statistics for these datasets. The number of isolates per species ranges from several hundreds for EC to 10 for PM.
These data were previously analyzed using PanX \cite{ding2018panx}, a pipeline for the analysis of bacterial pangenome. The pipeline takes as input a set of annotated genomes, and it clusters and aligns homologous genes. This allowed us to identify core genes (i.e. genes present in all isolates), and evaluate their average pairwise sequence divergence $\avg{d_\text{core}}$. We also evaluated the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$, the total core genome length $\Lcore$, and the total soft-core genome length $\Lsoftcore$, where soft-core genes are defined as genes being present in at least 95\% of the isolates. These quantities are reported in Table~I in the main text.\\
EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome. MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome fraction. The two datasets with highest sequence divergence are HP ($\dcore \sim 4\%$) and PM ($\dcore \sim 24\%$). For HP the average sequence divergence sits on the limit of what \textit{PanGraph} is able to accurately merge when using the \textit{minimap2} kernel (cf. Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel. The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.\\
We point out that the core genome size of the EC dataset found by PanX is significantly smaller than typical values seen in the literature, which are closer to 2 Mbp (see for example \cite{horsfield2023accurate}).\\
We point out that the value of $\Lcore = 0.7$ Mbp for EC is likely to be biased by annotation errors, to which the PanX pipeline is sensitive. Values for the core genome size of E.~coli reported in the literature are usually between 1500 and 2500 genes \cite{sutton2021pan,horsfield2023accurate,park2019large}, depending on the core genome definition used and the method used to mitigate annotation errors. For example using a strict definition for core genes the authors in \cite{sutton2021pan} initially find a core genome size of $\sim 700$ genes on a dataset of $\sim 1000$ isolates. However correcting for annotation errors this estimate increases to $\sim 1500$ genes. In our case we do not correct for these errors, since we only use core gene alignment to estimate the diversity of our dataset and since PanGraph is not sensitive to annotation errors. But we point out that for the EC dataset the value of $\Lsoftcore$ is likely to provide a better estimate of the core genome size.

>>>>

We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species: \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Prochlorococcus marinus} (PM), \textit{Mycobacterium tuberculosis} (MT) and \textit{Escherichia coli} (EC). In Table~I in the main text we report some summary statistics for these datasets. The number of isolates per species ranges from several hundreds for EC to 10 for PM.\\
These data has been chosen as part of a PanX \cite{ding2018panx} analysis dated March 2018.\footnote{
    We report for completeness that two of the isolates from the \textit{E.~coli} dataset from this analysis, namely \texttt{NC\_020518} and  \texttt{NC\_017663}, were subsequently removed from RefSeq. We verified that their removal from the dataset does not significantly impact core-genome and soft-core genome size, but decided to utilize the complete dataset as per original PanX analysis.
}
PanX is a pipeline for the analysis of bacterial pangenome. The pipeline takes as input a set of annotated genomes, and it clusters and aligns homologous genes. We used the PanX analysis to estimate the dataset diversity.\\
From this analysis we evaluated the average pairwise sequence divergence $\avg{d_\text{core}}$ of core genes, the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$, the total core genome length $\Lcore$, and the total soft-core genome length $\Lsoftcore$, where soft-core genes are defined as genes being present in at least 95\% of the isolates. These quantities are reported in Table~I in the main text.\footnote{
    $\Lcore$ is the length of the alignment used to estimate sequence divergence $\avg{d_\text{core}}$. However this quantity is sensitive to annotation errors. This is particularly evident in the \textit{E.~coli} dataset, for which $\Lcore = 0.7$ Mbp is likelt to be a subestimation of the real core genome size. $\Lsoftcore$ is instead less sensitive to annotation errors and usually considered a better core-genome size estimate. For example the \textit{E.~coli} core genome is usually considered to be aroud 1500-2500 genes in size, depending on the definition used \cite{park2019large,sutton2021pan,horsfield2023accurate}.
}\\
EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome. MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome fraction. The two datasets with highest sequence divergence are HP ($\dcore \sim 4\%$) and PM ($\dcore \sim 24\%$). For HP the average sequence divergence sits on the limit of what \textit{PanGraph} is able to accurately merge when using the \textit{minimap2} kernel (cf. Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel. The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.
